---
title: "Probability, Matrix Algebra"
author: "45tkg"
date: "12/22/2021"
output: 
  html_document:
    toc: true
    theme: united
header-includes: \usepackage{luatexja}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Random Variables

### Density function密度関数

Density functions represent realization patterns of continuous random variables. The following is the density function of the standard normal distribution:
$$\phi(x) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left\{- \frac{(x-\mu)^2}{2\sigma^2}\right\}$$

```{r normal-density}
x <- seq(-5, 5, length=100)
y <- dnorm(x, mean=0, sd=1)
plot(x, y, type="l")
```

The function **dnorm** means the density of a normal distribution. Similarly, **dt**, **dchisq**, and **dunif** are functions for the density of t, chi-squared, uniform distribution.

### Probability

The function **pnorm**

```{r probs}
p1 <- pnorm(2)                        # Pr[ X < 2 ]
p2 <- 1-pnorm(2)                      # Pr[ X > 2 ]
p3 <- pnorm(2, lower.tail=FALSE)      # Pr[ X > 2 ]
p4 <- pnorm(1)-pnorm(-2)              # Pr[ -2 < X < 1 ]
print(c(p1, p2, p3, p4))
```

```{r prob-norm, echo=FALSE}
x <- seq(-5, 5, length=100)
y <- dnorm(x, mean=0, sd=1)
plot(x,y,type="l")
xvals <- seq(-2,1,  length=20)    
dvals <- dnorm(xvals,mean=0,sd=1)
polygon(c(xvals,rev(xvals)),c(rep(0,20),rev(dvals)),col="red") 
arrows(-3.5, 0.2, -1.7, .12)
text(-3.5,0.22,"Pr[ -2.0 < Z < 1.0]")
#
xvals <- seq(2,5,  length=20)    
dvals <- dnorm(xvals, mean=0, sd=1)
polygon(c(xvals,rev(xvals)),c(rep(0,20),rev(dvals)),col="blue") 
arrows(3, 0.2, 2.5, .035)
text(3, 0.22, "Pr[ Z > 2.0]")
```

The function **qnorm** is quantile function for the normal distribution.

```{r}
q1 <- qnorm(0.95)               # { x : Pr[ X < x ] = 0.95}
print(q1)
```

```{r q-norm, echo=FALSE}
x <- seq(-5, 5, length=100)
y <- pnorm(x)
plot(x, y, type="l",lwd=2)
lines(c(-6,qnorm(.95)), c(.95,.95), lt=3, lwd=2.5,col="red")
lines(c(qnorm(.95),qnorm(.95)), c(.95,-1), lt=3, lwd=2.5,col="red")
arrows(3.0, 0.4, qnorm(.95)+0.1, .005)
text(3.5,0.48,"Pr[ Z < z ] = 0.95")
text(3.5,0.43,"( z = 1.64 )   ")
```


```{r t-both, echo=FALSE}
x <- seq(-5, 5, length=100)
y <- dt(x, df=3)
plot(x,y,type="l",ylim=c(0,0.4))
#xvals <- seq(-2,1,  length=20)    
#dvals <- dt(xvals,df=3)
#polygon(c(xvals,rev(xvals)),c(rep(0,20),rev(dvals)),col="red") 
#arrows(-3.5, 0.2, -1.7, .12)
#text(-3.5,0.22,"Pr[ -2.0 < X < 1.0]")
#
xvals <- seq(qt(0.95,df=3),5,  length=20)    
dvals <- dt(xvals,df=3)
polygon(c(xvals,rev(xvals)),c(rep(0,20),rev(dvals)),col="yellow") 
arrows(3.5, 0.18, qt(0.95,df=3), 0.0)
h1 <- expression(paste( Pr( Z < alpha ) == 0.95))
h2 <- expression(paste( Pr( Z > alpha ) == 0.05))
h3 <- expression(paste( ( alpha  == 1.64 ) ) )
text(3.5,0.26,h1)
text(3.5,0.23,h2)
text(3.5,0.20,h3)
```

## Other distributions

### t distribution

```{r t-prob-n}
pt(1, df=3) - pt(-2, df=3)        # Pr[ 1 < X < 3 ]
```

```{r t-prob-f, echo=FALSE}
x <- seq(-5, 5, length=100)
y <- dt(x, df=3)
plot(x, y, type="l")
xvals <- seq(-2, 1, length=20)    
dvals <- dt(xvals, df=3)
polygon(c(xvals,rev(xvals)),c(rep(0,20),rev(dvals)),col="red") 
arrows(-3.5, 0.2, -1.7, .12)
text(-3.5,0.22,"Pr[ -2.0 < T < 1.0 ]")
```

```{r t-q-n}
qt(0.60, df=3)
```

```{r t-q-f, echo=FALSE}
x <- seq(-5, 5, length=100)
y <- pt(x, df=3)
plot(x, y, type="l",lwd=2)
lines(c(-6,qt(.60, df=3)), c(.6,.6), lt=3, lwd=2.5,col="red")
lines(c(qt(.60, df=3),qt(.60, df=3)), c(.60,-1), lt=3, lwd=2.5,col="red")
arrows(3.0, 0.4, qt(.6,df=3)+0.1, .005)
text(2.5,0.48,"Pr[ T < t ] = 0.60")
text(2.5,0.42,"(t = 0.2766707)")
```

### Chi-squared distribution

```{r c-prob-n}
pchisq(3, df=5)-pchisq(1, df=5)   # Pr[ 1 < X < 3 ]
```

```{r c-prob-f, echo=FALSE}
x <- seq(0, 15, length=100)
y <- dchisq(x, df=5)
plot(x, y, type="l")
xvals <- seq(1, 3, length=20)    
dvals <- dchisq(xvals, df=5)
polygon(c(xvals,rev(xvals)),c(rep(0,20),rev(dvals)),col="red") 
arrows(9.5, 0.1, 3.2, 0.1)
text(12.0, 0.1, "Pr[ 1.0 < X < 3.0 ]")
```

```{r c-q-n}
qchisq(0.60, df=5)
```

```{r c-q-f, echo=FALSE}
x <- seq(0, 15, length=100)
y <- pchisq(x, df=5)
plot(x, y, type="l",lwd=2)
lines(c(-6,qchisq(.60, df=5)), c(.6,.6), lt=3, lwd=2.5,col="red")
lines(c(qchisq(.60, df=5),qchisq(.60, df=5)), c(.60,-1), lt=3, lwd=2.5,col="red")
arrows(10.0, 0.4, qchisq(.6, df=5)+0.1, .001)
text(10, 0.51,"Pr[ X < x ] = 0.60")
text(10, 0.45,"(x = 5.131867)")
```


### F distribution

```{r f-prob-n}
pf(3, df1=4, df2=10)-pf(1, df1=4, df2=10)
```

```{r f-prob-f, echo=FALSE}
x <- seq(0, 5, length=100)
y <- df(x, df1=10, df2=10)
plot(x, y, type="l")
xvals <- seq(1, 3, length=20)    
dvals <- df(xvals, df1=10, df2=10)
polygon(c(xvals,rev(xvals)),c(rep(0,20),rev(dvals)),col="red") 
arrows(3.5, 0.5, 2, .2)
text(3.5, 0.52, "Pr[ 1.0 < F < 3.0 ]")
```

```{r f-q-n}
qf(0.60, df1=4,df2=10)
```

```{r f-q-f, echo=FALSE}
x <- seq(0, 5, length=100)
y <- pf(x, df1=10, df2=10)
plot(x, y, type="l", lwd=2)
lines(c(-6,qf(.60, df1=10,df2=10)), c(.6,.6), lt=3, lwd=2.5,col="red")
lines(c(qf(.60, df1=10,df2=10),qf(.60, df1=10,df2=10)), c(.60,-1), lt=3, lwd=2.5,col="red")
arrows(3.0, 0.4, qf(.60, df1=10,df2=10)+0.1, .001)
text(3.5, 0.51,"Pr[ F < f ] = 0.60")
text(3.5, 0.45,"(f = 1.120273)")
```

### Bivariate Normal distribution
```{r bn-1}
binorm <- function(x, m, Sig){
  exp(-.5 * diag(t(x-m) %*% solve(Sig) %*% (x-m))) / sqrt(2 * pi * det(Sig))
}
m   <- c(0,0)
Sig0 <- matrix(c(1.0, 0.0, 0.0, 1.0), 2, 2)
Sig5 <- matrix(c(1.0, 0.5, 0.5, 1.0), 2, 2)
x  <- seq(-3, 3, length=50)
y  <- seq(-3, 3, length=50)
#### z ###############################################
fxy0      <- binorm( t(expand.grid(x,y)), m, Sig0)
dim(fxy0) <- c(length(x),length(y))
fxy5      <- binorm( t(expand.grid(x,y)), m, Sig5)
dim(fxy5) <- c(length(x),length(y))
###################################################################
#library(rgl)
#surface3d(x, y, fxy5, theta = 345, phi = 35,col="blue",front="lines")
```

```{r}
par(cex.lab=2)
par(cex.axis=1.75)
par(las=1)
par(mfrow=c(3,2))  # 1.5:1 aspect ratio

####################################################################
####################################################################
  ## Perspective Plot
  persp(x, y, fxy0, theta = 15, phi = 40,col="white",main="Bivariate Normal Density (corr=0.0)")
  persp(x, y, fxy5, theta = 15, phi = 40,col="white",main="Bivariate Normal Density (corr=0.5)")
  #mtext(expr.seq[j])

####################################################################
  ## Contour Plot
####################################################################
par(lwd=2,mgp=c(2,1,0))
contour(x, y, fxy0, col="blue", drawlabels=TRUE, nlevels=6,
        xlab=expression(X), ylab=expression(Y), 
        main="Contour / Conditional Density at X=0.7",
        lwd=2, xlim=range(x), ylim=range(y),
        frame.plot=TRUE, axes=FALSE, xaxs = "i", yaxs = "i")
axis(1, labels=TRUE, lwd.ticks=2)
axis(2, labels=TRUE, lwd.ticks=2)
##############################################
abline(v=.7,  lwd=1.5, lty=2, col="blue")
##############################################
```



```{r}
####################################################################
  ## Contour Plot
####################################################################
par(lwd=2,mgp=c(2,1,0))
contour(x, y, fxy5, col="blue", drawlabels=TRUE, nlevels=6,
        xlab=expression(X), ylab=expression(Y), 
        main="Contour / Conditional Density at X=0.7",
        lwd=2, xlim=range(x), ylim=range(y),
        frame.plot=TRUE, axes=FALSE, xaxs = "i", yaxs = "i")
axis(1, labels=TRUE, lwd.ticks=2)
axis(2, labels=TRUE, lwd.ticks=2)
##############################################
abline(v=.7,  lwd=1.5, lty=2, col="blue")
##############################################

###############################################################
f    <- function(x1,x2) binorm(t(cbind(x1,x2)), m, Sig5)
###############################################################
x.s  <- seq(from=min(x), to=max(x), by=0.1)
vals <- f(0.7, x.s)
fx <- dnorm(0.7, m[1], Sig5[1,1]^0.5)
lines( vals-abs(min(x)), x.s, lty=2, lwd=1.5, col="blue")
lines( vals/fx-abs(min(x)), x.s, lty=1, lwd=1.5, col="red")
###############################################################
text(1, -2, labels=expression(X==0.7))
text(-2.0, 1.5, labels=expression( f(y / x==0.7) ), col="red")
text(-2.2, 0.5, labels=expression( f(0.7, y) ),     col="blue")
legend(-2, 3, c("sectional view at x=0.7","condi. dens. of Y at x=0.7"),
lty=c(2,1),col=c("blue","red"),lwd=1.5)
```



```{r}
par(lwd=2,mgp=c(2,1,0))
contour(x, y, fxy0, col="blue", drawlabels=TRUE, nlevels=6,
        xlab=expression(X), ylab=expression(Y), 
        main="Contour / Marginal Density", 
        lwd=2, xlim=range(x), ylim=range(y),
        frame.plot=TRUE, axes=FALSE, xaxs = "i", yaxs = "i")
axis(1, labels=TRUE, lwd.ticks=2)
axis(2, labels=TRUE, lwd.ticks=2)
###############################################################
# Marginal density function: Left side solid
###############################################################
marginal.X <- function(x)  exp(-(x-m[1])^2/(2*Sig0[1,1])) / (sqrt(2*pi*Sig0[1,1]))
# Left side solid
vals <- marginal.X(x.s)
lines(vals-abs(min(x)), x.s, lty=1, lwd=2, col="red")
text(-2.3, 0.5, labels=expression( f(y) ), col="red")
legend(-2, 2.8, c("marginal denssity of Y"),lty=c(1),col=c("red"),lwd=2)
```


```{r}
par(lwd=2,mgp=c(2,1,0))
contour(x, y, fxy5, col="blue", drawlabels=TRUE, nlevels=6,
        xlab=expression(X), ylab=expression(Y), 
        main="Contour / Marginal Density", 
        lwd=2, xlim=range(x), ylim=range(y),
        frame.plot=TRUE, axes=FALSE, xaxs = "i", yaxs = "i")
axis(1, labels=TRUE, lwd.ticks=2)
axis(2, labels=TRUE, lwd.ticks=2)
###############################################################
# Marginal density function: Left side solid
###############################################################
marginal.X <- function(x)  exp(-(x-m[1])^2/(2*Sig5[1,1])) / (sqrt(2*pi*Sig5[1,1]))
# Left side solid
vals <- marginal.X(x.s)
lines(vals-abs(min(x)), x.s, lty=1, lwd=2, col="red")
text(-2.3, 0.5, labels=expression( f(y) ), col="red")
legend(-2, 2.8, c("marginal density of Y"),lty=c(1),col=c("red"),lwd=2)
```



## Matrix Algebra



